Illustrating Model Properties

### 2024-06-03 model.qmd
work on marginal over beta posteriors

```{python}
#| label: fig-model-prevalences-overall
#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension."
import matplotlib.pyplot as plt
from matplotlib import gridspec
import numpy as np
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS
from scripts import paths, shared

@@

# manually place legends
shared.split_legends(
  ax=axes[0,0],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.57, 0.55), (0.48, 0.99), (0.15, 0.99)],
)
shared.split_legends(
  ax=axes[0,1],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.67, 0.99), (0.27, 0.6), (0.1, 0.99)],
)
shared.split_legends(
  ax=axes[1,0],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.64, 0.99), (0.27, 0.6), (0.24, 0.99)],
)
shared.split_legends(
  ax=axes[1,1],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.64, 0.99), (0.4, 0.6), (0.16, 0.99)],
)

@@

```{python}
#| label: fig-model-prevalences-with-ipsi
#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown are four scenarios, all including contralateral LNL II involvement: Early T-category and an ipsilateral N0 neck (green), early T-category and ipsilateral LNL II involvement (blue), as well as the same two scenarios but for advanced T-category (orange and red). The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement. Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough."
import matplotlib.pyplot as plt
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS
from scripts import paths, shared

fig, ax = plt.subplots()
content = []
colors = iter([COLORS[c] for c in ["orange", "green", "red", "blue"]])

for t_stage in ["early", "late"]:
  for diagnosis in [{"II": True}, {lnl: False for lnl in ["I", "II", "III", "IV", "V"]}]:
    color = next(colors)
    scenario = Scenario(
      t_stages=[t_stage],
      diagnosis={
        "ipsi": {"max_llh": diagnosis},
        "contra": {"max_llh": {"II": True}},
      }
    )
    content.append(
      Histogram.from_hdf5(
        filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
        dataname=scenario.md5_hash("prevalences"),
        color=color,
        label="prediction"
      )
    )
    beta_post = BetaPosterior.from_hdf5(
      filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
      dataname=scenario.md5_hash("prevalences"),
      color=color,
    )
    beta_post.kwargs["label"] = f"data: {beta_post.num_success} of {beta_post.num_total}"
    content.append(beta_post)

draw(ax, content, xlims=(0., 25.))
ax.set_title("Observed vs Predicted Prevalence\nof Contralateral LNL II Involvement")
ax.set_yticks([])
ax.set_xlabel("prevalence [%]")

shared.split_legends(
  ax=ax,
  titles=["early with ipsi II", "early ipsi N0", "late with ipsi II", "late with ipsi N0"],
  locs=[(0.23, 0.65), (0.1, 0.99), (0.6, 0.65), (0.45, 0.99)],
)

@@

```{python}
#| label: fig-model-prevalences-midext
#| fig-cap: "Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category. While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data. Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue."
import matplotlib.pyplot as plt
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS
from scripts import paths, shared

@@

for t_stage in ["early", "late"]:
  color = next(colors)
  content.append(
    Histogram.from_hdf5(
      filename=paths.model_dir / "full" / "prevalences_midext.hdf5",
      dataname=t_stage,
      color=color,
      label="prediction"
    )
  )
  beta_post = BetaPosterior.from_hdf5(
    filename=paths.model_dir / "full" / "prevalences_midext.hdf5",
    dataname=t_stage,
    color=color,
  )
  beta_post.kwargs["label"] = f"data: {beta_post.num_success} of {beta_post.num_total}"
  content.append(beta_post)

@@

shared.split_legends(
  ax=ax,
  titles=["early T-cat.", "late T-cat."],
  locs=[(0.05, 0.99), (0.6, 0.99)],
)

>>>>

```{python}
#| label: fig-model-prevalences-overall
#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension."
import matplotlib.pyplot as plt
from matplotlib import gridspec
import numpy as np
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS, split_legends
from scripts import paths, shared

@@

# manually place legends
split_legends(
  axes=axes[0,0],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.57, 0.55), (0.48, 0.99), (0.15, 0.99)],
)
split_legends(
  axes=axes[0,1],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.67, 0.99), (0.27, 0.6), (0.1, 0.99)],
)
split_legends(
  axes=axes[1,0],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.64, 0.99), (0.27, 0.6), (0.24, 0.99)],
)
split_legends(
  axes=axes[1,1],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.64, 0.99), (0.4, 0.6), (0.16, 0.99)],
)

@@

```{python}
#| label: fig-model-prevalences-with-ipsi
#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown are four scenarios, all including contralateral LNL II involvement: Early T-category and an ipsilateral N0 neck (green), early T-category and ipsilateral LNL II involvement (blue), as well as the same two scenarios but for advanced T-category (orange and red). The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement. Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough."
import matplotlib.pyplot as plt
from matplotlib import gridspec
import numpy as np
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS, split_legends
from scripts import paths, shared
from scripts.marginalize_midext import (
  compute_possible_observed_prevalences,
  SummedBetaPosterior,
)

full.set_title(
  "Observed vs Predicted Prevalence of Contralateral LNL II Involvement\nDependent on Ipsilateral Involvement",
  pad=20.,
  fontsize="large",
)

data = shared.get_data()

for i, t_stage in enumerate(["early", "late"]):
  axes[i,0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
  for j, midext in enumerate([False, True]):
    axes[0,j].set_title("Mid. ext." if midext else "Lateralized", fontweight="bold")
    axes[0,j].set_xticks([])
    axes[-1,j].set_xlabel("prevalence [%]")
    axes[i,j].set_yticks([])
    content = []
    for ipsi, color in zip(
      [{"I": False, "II": False, "III": False, "IV": False, "V": False}, {"III": True}],
      [COLORS["blue"], COLORS["orange"]],
    ):
      scenario = Scenario(
        t_stages=[t_stage],
        midext=midext,
        diagnosis={
          "ipsi": {"max_llh": ipsi},
          "contra": {"max_llh": {"II": True}},
        }
      )
      content.append(
        Histogram.from_hdf5(
          filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
          dataname=scenario.md5_hash("prevalences"),
          color=color,
          label="prediction",
        )
      )
      content.append(
        BetaPosterior.from_hdf5(
          filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
          dataname=scenario.md5_hash("prevalences"),
          color=color,
          label="beta",
        )
      )
      num_success, num_total = compute_possible_observed_prevalences(data, scenario).T
      content.append(
          SummedBetaPosterior(
          num_success=num_success,
          num_total=num_total,
          kwargs={"color": "red"},
        )
      )
    draw(
      axes=axes[i,j],
      contents=content,
      xlims=(0., 30. if midext else 10.),
    )

# manually place legends
split_legends(
  axes=axes[0,0],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.3, 0.99), (0.45, 0.55)],
)
split_legends(
  axes=axes[0,1],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.3, 0.99), (0.5, 0.55)],
)
split_legends(
  axes=axes[1,0],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.2, 0.99), (0.55, 0.99)],
)
split_legends(
  axes=axes[1,1],
  titles=["ipsi N0", "ipsi LNL III"],
  locs=[(0.2, 0.99), (0.65, 0.99)],
)

@@

```{python}
#| label: fig-model-prevalences-midext
#| fig-cap: "Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category. While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data. Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue."
import matplotlib.pyplot as plt
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS, split_legends
from scripts import paths, shared
from scripts.marginalize_midext import compute_match_total_combinations, SummedBetaPosterior

@@

data = shared.get_data()
has_unknown_midext = data.ly.midext.isna()

for t_stage in ["early", "late"]:
  has_t_stage = data.ly.t_stage.isin([t_stage])
  scenario = Scenario(
    t_stages=[t_stage],
    midext=True,
    diagnosis={"ipsi": {"max_llh": {"II": None}}},
  )
  color = next(colors)
  content.append(
    Histogram.from_hdf5(
      filename=paths.model_dir / "full" / "prevalences_midext.hdf5",
      dataname=t_stage,
      color=color,
      label="prediction"
    )
  )
  beta_post = BetaPosterior.from_hdf5(
    filename=paths.model_dir / "full" / "prevalences_midext.hdf5",
    dataname=t_stage,
    color=color,
  )
  base_match = beta_post.num_success
  base_total = beta_post.num_total
  num_success, num_total = compute_match_total_combinations(
    base_match=base_match,
    base_total=base_total,
    unknown_match=(has_unknown_midext & has_t_stage).sum(),
    unknown_total=(has_unknown_midext & has_t_stage).sum(),
  ).T
  content.append(
    SummedBetaPosterior(
      num_success=num_success,
      num_total=num_total,
    )
  )

@@

split_legends(
  axes=ax,
  titles=["early T-cat.", "late T-cat."],
  locs=[(0.05, 0.99), (0.6, 0.99)],
)
